NumPy---Array Computation for Python

### 2020-01-10 body.tex
Revise intro

The first incarnation of NumPy was a package called Numeric, written in 1995
by Jim Hugunin, then completing his master's thesis at MIT.
Hugunin based his package on previous work by Jim Fulton,
then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Later, Paul Dubois at Lawrence Livermore National Laboratory toke over the project
and funded the writing of the first
manual\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.
% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
% May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)
% https://aip.scitation.org/doi/pdf/10.1063/1.4823451 
% https://aip.scitation.org/doi/pdf/10.1063/1.4822400

% S TODO: the first sentence below could flow a bit better
Around 2000, the Space Telescope Science Institute (STScI) software group wrote
a re-implementation of much of Numeric, called NumArray, to support their
need to work on large memory-mapped arrays and arrays of mixed data type
records \cite{STScI-slither}.
This briefly caused the Numeric and NumArray communities to diverge, until 2005,
when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
This merger resulted in NumPy.

NumPy now underpins almost every Python library that does scientific or
numerical computation, including SciPy\cite{virtanen2019scipy},
matplotlib\cite{hunter2007matplotlib}, pandas\cite{mckinney-proc-scipy-2010},
scikit-learn\cite{pedregosa2011scikit}, and
scikit-image\cite{vanderwalt2014scikit}.
NumPy consists of the $n$-dimensional array object along with utility functions
that operate on it.
Because of its inherent simplicity---being a pointer to memory with some
associated meta-data about shape, data-type, and so forth---the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for tensor computation---witnessed by its use in popular
deep learning libraries such as PyTorch\cite{pytorch}.

@@

The NumPy array is a structure that efficiently stores and accesses
$n$-dimensional array data\cite{vanderwalt2011numpy}.
The need for such a data structure arises often in science.
For instance, it may be used to store experimental measurements made at
discrete time intervals (1-D arrays), gray-level photographs (2-D arrays),
spectral measurement on a regular grid (3-D arrays), or color videos (4-D
arrays).

The NumPy array consists of a pointer to memory, along with meta-data used to
interpret the data stored there.
Computer memory stores elements linearly and has no notion
of higher-dimensions, but we can interpret it as an $n$-dimensional array by also
providing the array's shape, strides (the number of bytes to advance in memory to skip
along rows, columns, etc.), and data-type (see Figure~\ref{fig:array-concepts}).

Data types (dtypes for short) tell us more about the elements stored in a NumPy
array.
Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
timestamps, or Python objects, each identified by its own dtype.

@@

As Python's popularity for scientific computing, data science, and machine
learning increased, many new use cases arose.

>>>>

Two different Python array packages existed before NumPy.
The Numeric package began in the mid-90s and provided an array object in
Python, written in C, and linking to standard fast implementations of linear
algebra.
Around 2000, the Space Telescope Science Institute (STScI) software group wrote
a re-implementation of much of Numeric, called NumArray, to support their
need to work on large memory-mapped arrays and arrays of mixed data type
records \cite{STScI-slither}.
This briefly caused the two communities to diverge, until
2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
and NumArray \cite{oliphant2006guide}.

%By the beginning of the millennium, several libraries built on top of Numeric 
%to provide fundamental scientific routines were combined into SciPy.
%Due to the poor performance of Numeric on large data sets, NumArray was
%developed at the Space Telescope Science Institute (STScI) to handle
%large astronomical images such as those produced by the Hubble Space Telescope.
%However, while initially considered the successor to Numeric, NumArray was no longer
%performant on multiple small arrays.
%This divide posed a challenge for SciPy as it wasn't clear how to support
%both libraries.
%Working with the NumArray and SciPy communities, the then maintainer of Numeric,
%Travis Oliphant, embarked upon a major rewrite that aimed to be a ``best of both
%worlds'' unification of Numeric and NumArray. This merger resulted in NumPy.
%
%The first incarnation of NumPy was a package called Numeric, written in 1995
%by Jim Hugunin, then completing his master's thesis at MIT.
%Hugunin based his package on previous work by Jim Fulton,
%then working at the US Geological Survey, and acknowledges
%many others in his initial report of the package \cite{Hugunin-whitepaper}.
%Later, Paul Dubois at Lawrence Livermore National Laboratory toke over the project
%and funded the writing of the first
%manual\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
%Numeric provided an array object in Python, written in C, and linking to
%standard fast implementations of linear algebra.
%% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
%% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
%% May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)
%% https://aip.scitation.org/doi/pdf/10.1063/1.4823451 
%% https://aip.scitation.org/doi/pdf/10.1063/1.4822400
%
%% S TODO: the first sentence below could flow a bit better
%Around 2000, the Space Telescope Science Institute (STScI) software group wrote
%a re-implementation of much of Numeric, called NumArray, to support their
%need to work on large memory-mapped arrays and arrays of mixed data type
%records \cite{STScI-slither}.
%This briefly caused the Numeric and NumArray communities to diverge, until 2005,
%when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
%both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
%This merger resulted in NumPy.

Today, NumPy underpins almost every Python library that does scientific or
numerical computation including SciPy\cite{virtanen2019scipy},
matplotlib\cite{hunter2007matplotlib}, pandas\cite{mckinney-proc-scipy-2010},
scikit-learn\cite{pedregosa2011scikit}, and
scikit-image\cite{vanderwalt2014scikit}.
It consists of an $n$-dimensional array object along with utility functions
that operate on it.
Because of its inherent simplicity---being a pointer to memory with some
associated meta-data---the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for array computation.

@@

The NumPy array is a structure that efficiently stores and accesses
$n$-dimensional array data\cite{vanderwalt2011numpy}.
The need for such a data structure arises often in science.
For instance, it may be used to store experimental measurements made at
discrete time intervals (1-D arrays), gray-level photographs (2-D arrays),
spectral measurements on a regular grid (3-D arrays), or color videos (4-D
arrays).

The NumPy array consists of a pointer to memory, along with meta-data used to
interpret the data stored there.
Computer memory stores elements linearly and has no notion
of higher-dimensions, but we can interpret it as an $n$-dimensional array by also
providing the array's shape, strides (the number of bytes to advance in memory to skip
along rows, columns, etc.), and data type (see Figure~\ref{fig:array-concepts}).

Data types (dtypes for short) tell us more about the elements stored in a NumPy
array.
% can hold ...?, e.g.,
Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
timestamps, or Python objects, each identified by its own dtype.

@@

As Python's popularity for scientific computing, data science, and machine
learning increased, many new challenges and use cases arose.
